---
title: "Data Management With R: Exploratory Data Analysis"
author: "Matthias Haber"
date: "25 September 2017"
output:
  beamer_presentation:
    theme: "metropolis"
    colortheme: "default"
    fonttheme: "default"
    fig_caption: false
    df_print: default
    toc: true
  ioslides_presentation:
  slidy_presentation: default
  linkcolor: "red"
---

# Last week's homework

# Prerequisites

## Packages

```{r, message=FALSE}
library(tidyverse)
```

## Data

336,776 flights that departed from New York City in 2013

```{r, message=FALSE}
# install.packages("nycflights13")
library(nycflights13)
```

```{r echo = FALSE}
knitr::kable(flights[1:4, 1:6])
```

# Exploratory Data Analysis

## Goals

> "There are no routine statistical questions, only questionable statistical
> routines." --- Sir David Cox

> "Far better an approximate answer to the right question, which is often
> vague, than an exact answer to the wrong question, which can always be made
> precise." --- John Tukey

Learn how to use visualization and transformation to systematically explore your data to answer or generate questions about your data.


# Homework Exercises

## Homework Exercises
For this week's homework exersises go to Moodle and answer the Quiz posted in the Data Transformation section. You will be asked a number of questions randomly selected from a question pool. If you work in pairs, then you might get two different sets of questions. 

Deadline: Sunday, October 1 before midnight.

##  {.standout}
That's it for today. Questions?    

